---
title: "Introduction to birdscanR"
author: Birgen Haest, Baptiste Schmid, Fabian Hertner
output: rmarkdown::html_vignette
description: > 
  Which functions do I need to call to calculate migration traffic rates from data stored in an SQL Birdscan database? This vignette describes the main workflow to do so.
vignette: >
  %\VignetteIndexEntry{Introduction to birdscanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
``` {r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```
***
# Summary

This vignette provides an example of the basic workflow to go from an 'MS-SQL' or 'PostgreSQL' Birdscan database to Activity Rates, or Migration Traffic Rates for the target animal classes. The workflow includes 4 steps, of which step 2 is optional, i.e., only of interest when you also want to know about possible migratory bat activity):

1.    Extract the relevant data from the SQL database.

2.    (optional) Join the bat classification with the base MR1 classification output (that contains insects and different bird groups). This step is only needed if you want to calculate migration traffic rates for bats, as the bat classification is not part of the standard MR1 classification output. If you are only interested in birds and insects, you can skip this step.

3.    Filter the data to your target output period, classes, and other preferences.

4.    Calculate the activity or migration traffic rates

***
# Before we get started

We first load the birdscanR package.
```{r}
library(birdscanR)
```

Then we set our input:
```{r, eval=FALSE}
# Set main output directory
# =============================================================================
mainOutputDir = file.path(".", "results")

# Set server and database settings
# =============================================================================
dbServer = "MACHINE\\SERVERNAME" # Set the name of your SQL server
dbName = "db_Name" # Set the name of your database
dbDriverChar = "SQL Server" # Set either "SQL Server" or "PostgreSQL"

# Set timezone used by the radar/database
# Use "Etc/GMT0" (UTC) as radarTimeZone for birdscan v1.6
# and greater as all times are in UTC as of this software version.
# =============================================================================
radarTimeZone = "Etc/GMT0"

# Set target timezone for the radar/dataset
# Example: targetTimeZone = "Etc/GMT-1" (Etc/GMT-1 equals UTC+1)
# =============================================================================
targetTimeZone = "Etc/GMT0"

# Set list of Rf features you also want to extract
# Vector with RF features to extract. Feature IDs can be found in the
# rffeatures table in the sql database.
# Example: Get wing beat frequency and credibility: c(167, 168)
# Set to NULL to not extract any.
# =============================================================================
listOfRfFeaturesToExtract = NULL

# Geographic location of the radar measurements - c(Latitude, Longitude)
# =============================================================================
siteLocation = c(47.494427, 8.716432)

# Set type of twilight to use for day/night decision, i.e., "sun" or "civil"
# =============================================================================
sunOrCivil = "civil"

# Filter settings
# =============================================================================
# Set desired pulseLength modes
pulseLengthSelection = c("S")

# Set desired rotation modes (multiple simultaneous selections possible)
# options: 1 (rotation), 0 (nonrotation)
rotationSelection = c(1)

# Set classes to use in the analysis
# Example: classSelection = c("passerine_type", "wader_type", "swift_type",
#                             "large_bird", "unid_bird", "bird_flock",
#                             "insect", "nonbio", "precipitation")
classSelection = c("insect", "bat", "passerine_type")

# Set the classification propability cutoff (between 0 and 1) - only objects
# for which the chosen label also has a classification probability > than the
# cutoff are retained; Set to NULL to not subset according to probability.
classProbCutoff = NULL

# Set the altitude range (in meters agl)
altitudeRange = c(50, 1000)

# Set the time range for echodata (in the targetTimeZone)
# use format "yyyy-MM-dd hh:mm"
timeRangeData = c("2021-01-15 00:00", "2021-01-31 00:00")

# Set whether to get the manual blind time from:
# "mandb": the db file;
# "csv"  : a separate csv file
manBlindSource = "csv"

# Set paths to manual blind times file, if manBlindSource == "csv"
if (manBlindSource %in% "csv") {
  manblindFile = file.path("data", "manualBlindTimes.csv")
}

# Set whether to use the echoValidator - If set to TRUE, echoes labelled
# by the echo validator as “non-bio scatterer” will be excluded.
useEchoValidator = FALSE

# MTR calculation settings
# =============================================================================
# Set whether to save the blind times to file
saveBlindTimes = TRUE

# Set altitude Range and Bin Size for the MTR calculations
altitudeRange.mtr = c(50, 500)
altitudeBinSize = 50

# time range for timeBins (targetTimeZone) - format: "yyyy-MM-dd hh:mm"
timeRangesTimeBins = c("2021-01-15 00:00", "2021-01-31 00:00")

# timeBin size in seconds
timeBinduration_sec = 3600

# set blindtime types which should not be treated as blindtime but MTR = 0
blindTimeAsMtrZero = c("rain")

# cutoff for proportional observation times: Ignore TimeBins where
# "observationTime/timeBinDuration < propObsTimeCutoff"
# in 'computeMTR' only used if parameter 'computePerDayNight' is set to
# 'TRUE' and timeBins are shorter than day/night. In this case timeBins
# are combined by day/night and only time bins with a proportional
# observation time greater than the cutoff will be used to compute the
# day/night MTR and spread.
# set value from 0-1
propObsTimeCutoff = 0.2

# Set classes for which you want the MTR
# Example: classSelection = c("passerine_type", "wader_type", "large_bird")
# classSelection = c("passerine_type", "wader_type", "swift_type",
#                    "large_bird", "unid_bird", "bird_flock",
#                    "insect", "nonbio", "precipitation")
classSelection.mtr = c("insect", "bat")

# Set whether to compute MTR per timebin or per day/night
# TRUE: MTR is computed per day and night;
# FALSE: MTR is computed for each time bin
computePerDayNight = FALSE

# Set whether to save the MTR to file
saveMTR2File = TRUE
```

***
# Step 1: Extract the relevant data from the SQL database (or load it from a )

Extracting the data that is need for MTR calculations, can be done with the `extractDbData()` function. This function will connect to your SQL database and read all the necessary data to memory. If requested by setting the parameter `saveDbToFile` to `TRUE` , it will also write it to an .rds file in the output directory you chose. The file is automatically named to: `[yourDbName]_DataExtract.rds`, and written to the main output directory you defined in the input settings.

If you currently have no connection to an SQL server with your MR1 database attached, and have extracted or received an extracted dataset earlier, you can directly load your database from the the RDS file, using 'dbData = readRDS(file.path(`directory containing the file`, `[yourDbName]_DataExtract.rds`))'.

```{r, eval=FALSE}
# Print progress message
# =============================================================================
message(paste0("Extracting data from ", dbName))

# Extract the data from the SQL database
# =============================================================================
dbData = extractDbData(
  dbDriverChar = dbDriverChar,
  dbServer = dbServer,
  dbName = dbName,
  saveDbToFile = TRUE,
  dbDataDir = mainOutputDir,
  radarTimeZone = radarTimeZone,
  targetTimeZone = targetTimeZone,
  listOfRfFeaturesToExtract = listOfRfFeaturesToExtract,
  siteLocation = siteLocation,
  sunOrCivil = sunOrCivil
)

# Print progress message
# =============================================================================
message(paste0("Finished extracting data from ", dbName))
```

***
# Step 2: (optional) Join the bat classification with the base MR1 classification output

In addition to the standard echo classification in insects, different bird groups, and non-biological echoes, Birdscan MR1 databases that were measuring with or have been updated to version 1.7.0.0 or higher, offer the possibility to classify echoes as bat.

The bat infromation is, however, included in separate tables in the SQL database. With `reclassToBats()` one can merge the bat classification information with the standard classification output. If you are not interested in bats, you can simply skip this step.

```{r, eval=FALSE}
# Join bat classification information with standard echo classifcation
# ===========================================================================
dbData$echoData = reclassToBats(
  echoData = dbData$echoData,
  batProbabilitiesAndMtrFactors = dbData$batProbabilitiesAndMtrFactors,
  reclassToBatCutoff = 0.5
)
```

***
# Step 3: Filter the extracted data to what you are interested in

Filtering both the protocol and echo data (extracted in the previous step) can be done in one step with the function `filterData()`.   

```{r, eval=FALSE}
# Print progress message
# =============================================================================
message(paste0("Filtering data from ", dbName))

# Get current manual blind times
# =============================================================================
# CASE: manBlindSource == "csv"
# ===========================================================================
if (manBlindSource %in% "csv") {
  # Read manual blindtimes from csv File
  # csv contains one row per blindtime and 3 columns
  # (start of blindtime, stop of blindtime, type of blindtime)
  # times have to be of format 'yyyy-MM-dd hh:mm:ss'
  # =======================================================================
  cManualBlindTimes = loadManualBlindTimes(
    filePath = manblindFile,
    blindTimesTZ = radarTimeZone,
    targetTZ = targetTimeZone
  )
  # CASE: manBlindSource == "mandb"
  # ===========================================================================
} else if (manBlindSource %in% "mandb") {
  cManualBlindTimes = dbData$manualVisibilityTable
}

# Filter the data
# =============================================================================
filteredEchoProtocol = filterData(
  echoData = dbData$echoData,
  protocolData = dbData$protocolData,
  pulseTypeSelection = pulseLengthSelection,
  rotationSelection = rotationSelection,
  timeRangeTargetTZ = timeRangeData,
  targetTimeZone = targetTimeZone,
  classSelection = classSelection,
  classProbCutOff = classProbCutoff,
  altitudeRange_AGL = altitudeRange,
  manualBlindTimes = cManualBlindTimes,
  echoValidator = useEchoValidator
)

# Save the filtered echo and protocol data to the database data list
# =============================================================================
dbData$echoData = filteredEchoProtocol$echoData
dbData$protocolData = filteredEchoProtocol$protocolData

# Save the filtered dataset to a file, including also all filter settings,
# and the other tables in the original dataset
# =============================================================================
dbData$echoFiltersApplied = list(
  classProbCutoff = classProbCutoff,
  altitudeRange_AGL = altitudeRange,
  targetTimeZone = targetTimeZone,
  timeRangeEchoData = timeRangeData,
  useEchoValidator = useEchoValidator
)
if (is.null(classProbCutoff)) {
  classProbCutoff.char = 0
} else {
  classProbCutoff.char = classProbCutoff
}
outputFile = file.path(
  mainOutputDir,
  paste0(
    dbName, "_filtered_",
    "cut", classProbCutoff.char, "_",
    "altRange", paste(altitudeRange, collapse = "to"), "_",
    "timeRange", paste(format(as.Date(timeRangeData), "%Y%m%d"),
      collapse = "to"
    ),
    "_",
    "echoVal", as.character(useEchoValidator),
    ".rds"
  )
)
saveRDS(dbData, outputFile)

# Print progress message
# =============================================================================
message(paste0("Finished filtering data from ", dbName))
```

# Step 4: Calculate the activity or migration traffic rates

Calculation of the migration traffic rates is done with the function computeMTR().

```{r, eval=FALSE}
# Print information message
# =====================================================================
if (computePerDayNight) {
  message(paste0(
    "Computing MTR for ", dbName, ", using:\n",
    "Classes: ", paste(classSelection.mtr, collapse = ", "), "\n",
    "For altitudes: ", paste(altitudeRange.mtr, collapse = " to "),
    " in bins of ", altitudeBinSize, "m\n",
    "on a nightly/daily basis"
  ))
} else {
  message(paste0(
    "Computing MTR for ", dbName, ", using:\n",
    "Classes: ", paste(classSelection.mtr, collapse = ", "), "\n",
    "For altitudes: ", paste(altitudeRange.mtr, collapse = " to "),
    " in bins of ", altitudeBinSize, "m\n",
    "Timebins of: ", timeBinduration_sec, " seconds or ",
    timeBinduration_sec / 3600, " hours"
  ))
}

# Calculate the MTR
# =====================================================================
mtr = computeMTR(
  dbName = dbName,
  echoes = dbData$echoData,
  classSelection = classSelection.mtr,
  altitudeRange = altitudeRange.mtr,
  altitudeBinSize = altitudeBinSize,
  timeRange = timeRangesTimeBins,
  timeBinDuration_sec = timeBinduration_sec,
  timeZone = targetTimeZone,
  sunriseSunset = dbData$sunriseSunset,
  sunOrCivil = sunOrCivil,
  protocolData = dbData$protocolData,
  visibilityData = dbData$visibilityData,
  manualBlindTimes = cManualBlindTimes,
  saveBlindTimes = saveBlindTimes,
  blindTimesOutputDir = mainOutputDir,
  blindTimeAsMtrZero = blindTimeAsMtrZero,
  propObsTimeCutoff = propObsTimeCutoff,
  computePerDayNight = computePerDayNight,
  computeAltitudeDistribution = TRUE
)

# Save the mTR to file, if requested
# =====================================================================
if (saveMTR2File) {
  if (computePerDayNight) {
    outputFile = paste0(
      "mtr_", dbName,
      "_alt", paste(altitudeRange.mtr,
        collapse = "to"
      ),
      "per", altitudeBinSize, "m_",
      "time",
      paste(format(as.Date(timeRangesTimeBins), "%Y%m%d"),
        collapse = "to"
      ),
      "perDayNight_",
      "cut", propObsTimeCutoff, ".rds"
    )
  } else {
    outputFile = paste0(
      "mtr_", dbName,
      "_alt", paste(altitudeRange.mtr,
        collapse = "to"
      ),
      "per", altitudeBinSize, "m_",
      "time", paste(format(as.Date(timeRangesTimeBins), "%Y%m%d"),
        collapse = "to"
      ),
      "per", timeBinduration_sec, "s_",
      "cut", propObsTimeCutoff, ".rds"
    )
  }
  saveMTR(
    mtr = mtr,
    filepath = mainOutputDir,
    fileName = outputFile
  )
}
```

# Create some plots to explore the calculated activity / migration traffic rates

The `birdscanR` package also provides two basic plotting functions to visualize and explore the MTR results. 

The function `plotLongitudinalMTR()` creates a time series plot of the MTR values (one per altitude bin in the MTR data):

```{r, eval=FALSE}
# Make time series plot
# =============================================================================
# Set time range for plots (in targetTimeZone) ;
# A plot is created for each timerange
# use format "yyyy-MM-dd hh:mm"
# ===========================================================================
timeRangePlot = createTimeRangeForPlot(timeRangeData[1], timeRangeData[2], 7)

# Set output path for plots
# ===========================================================================
outputDir.plots = file.path(mainOutputDir, "Plots")

# Set the class of which the MTR data should be plotted.
# If not set or set to “allClasses”, MTR of all classes will be plotted.
# ===========================================================================
plotClass = "passerine_type"

# Set the maximum value of the y-Scale of the plot to the given value.
# If negative or not set, the y-Scale is auto-scaled.
# ===========================================================================
maxMTR.plot = -1

# Set the propObsTimeCutOff for the plot
# Time bins with a proportional observation time smaller than
# propObsTimeCutoff will not be shown in the plot
# ===========================================================================
propObsTimeCutoff.plot = 0.2

# Set if the spread (first and third quartile) should be plotted
# ===========================================================================
plotSpread = TRUE

# Print message
# ===========================================================================
message("Plotting time series of MTR values..")

# Make Plot
# ===========================================================================
plotLongitudinalMTR(
  mtr = mtr,
  maxMTR = maxMTR.plot,
  timeRange = timeRangePlot,
  targetTimeZone = "Etc/GMT0",
  plotClass = plotClass,
  propObsTimeCutoff = propObsTimeCutoff.plot,
  plotSpread = plotSpread,
  filePath = outputDir.plots
)
```

With the second plotting function `plotExploration()` you can create a time series plot of the detected objects. Each object is shown at the respective height (m above ground level) and time:

```{r, eval=FALSE}
# Make an exploration plot
# =============================================================================
# Set  the maximum value of the y-Scale of the plot to the given value.
# If negative or not set, the y-Scale is auto-scaled.
maxAltitude.plot = -1

# Print message
# ===========================================================================
message("Plotting exploration..")

# Make Plot
# ===========================================================================
plotExploration(
  echoData = dbData$echoData,
  timeRange = timeRangePlot,
  targetTimeZone = "Etc/GMT0",
  manualBlindTimes = cManualBlindTimes,
  visibilityData = dbData$visibilityData,
  protocolData = dbData$protocolData,
  sunriseSunset = dbData$sunriseSunset,
  maxAltitude = maxAltitude.plot,
  filePath = outputDir.plots
)
```
