---
title: "Guide to extracting specific tables from a Birdscan MR1 SQL database"
author: Birgen Haest, Fabian Hertner, Baptiste Schmid
output: rmarkdown::html_vignette
description: > 
  Which functions do I need to call to get the data from specific tables from the Birdscsan MR1 SQL database? This vignette describes the main workflow to do so.
vignette: >
  %\VignetteIndexEntry{Guide to extracting specific tables from a Birdscan MR1 SQL database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
``` {r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```
***
# Summary
Instead of using `extractDbData()` to extract all relevant data at once to calculate densities or Migration Traffic Rates, one sometimes just wants to have a look at certain specific tables in the database like the `site`, `radar` or any of the other tables in the 'Birdscan MR1' 'SQL' database. This vignette provides some examples of how the different 'get' functions can be used to extract these different tables: 

- `getBatClassification()`: extract the bat classification table

- `getCollectionTable()`: extract the echo collection table

- `getEchoFeatures()`: extract the echo features table

- `getEchoValidationTable()`: extract the echo validation table

- `getProtocolTable()`: extract the protocol table

- `getRadarTable()`: extract the radar table

- `getRfClassification()`: extract the RF classification table

- `getSiteTable()`: extract the site table

- `getTimeBinsTable()`: extract the time bins table (note: These are not the time bins used in `computeMTR()` and `computeDensity()`)

- `getVisibilityTable()`: extract the visibility table


***
# Before we get started

We first load the birdscanR package:
```{r}
library(birdscanR)
```

Then we set our inputs:
```{r}
# Set main output directory
# =============================================================================
mainOutputDir = file.path(".", "results")
```

Open the connection to the SQL database: 
```{r, eval=FALSE}
# Set server and database settings
# =============================================================================
dbServer = "MACHINE\\SERVERNAME" # Set the name of your SQL server
dbName = "db_Name" # Set the name of your database
dbDriverChar = "SQL Server" # Set either "SQL Server" or "PostgreSQL"

# Open the connection with the database
# ===========================================================================
dsn = paste0(
  "driver=", dbDriverChar, ";server=", dbServer,
  ";database=", dbName,
  ";uid=", rstudioapi::askForPassword("Database user"),
  ";pwd=", rstudioapi::askForPassword("Database password")
)
dbConnection = RODBC::odbcDriverConnect(dsn)
```

***
# Examples

## Extracting the bat classification table

Use `getBatClassification()` to extract the table with the bat vs nonbat classifications.   
```{r, eval=FALSE}
# load collection table
# =============================================================================
message("Extracting the bat classification table from DB...")
batClassifications = getBatClassification(dbConnection, timeInterval)
```


## Extracting the echo collection table

Use `getCollectionTable()` to extract the echo data from the collection table.  The parameter `timeInterval` can be used to restrict extraction to a specific time interval only. 
```{r, eval=FALSE}
# load collection table
# =============================================================================
message("Extracting collection table from DB...")
timeInterval = NULL # e.g. c(as.Date("2024-03-01"),as.Date("2024-03-15"))
echoData = getCollectionTable(dbConnection, timeInterval)
```


## Extracting specific echo features form the echo_rffeature_map table

Use `getEchoFeatures()` to extract additional features stored for each sample.  For a full list of available features, and the corresponding feature numbers, see the `rffeatures` table in the SQL dataset. There are currently 190 features available. Use `c(0, 189)` to extract all features from the `echo_rffeature_map` table.
```{r, eval=FALSE}
# Example: Extract the wing beat frequency and credibility
# =============================================================================
message("Extracting rffeatures table from DB...")
listOfRfFeaturesToExtract = c(167, 168)
echoRfFeatureMap = getEchoFeatures(dbConnection, dbDriverChar,
  listOfRfFeaturesToExtract = listOfRfFeaturesToExtract
)

# Add the newly extracted features to the existing echoData
# =============================================================================
echoData %>% left_join(echoData, echoRfFeatureMap, by = join_by(row == echo))
```

## Extracting information on the study site

Use `getSiteTable()` to extract the `site` table that holds information about the site where the data was collected like the start and end dates of the radar campaign, latitude, longitude, altitude, and other details.
```{r, eval=FALSE}
# Extract the site  table from the SQL database
# =============================================================================
message("Extracting site table from DB...")
siteTable = getSiteTable(dbConnection)
```
***
