---
title: "How to extract single tables?"
author: Birgen Haest, Fabian Hertner, Baptiste Schmid
output: rmarkdown::html_vignette
description: > 
  Which functions do I need to call to get data from single tables? This vignette describes the main workflow to do so.
vignette: >
  %\VignetteIndexEntry{Introduction to birdscanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
``` {r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```
***
# Summary

This vignette details the function `extractDbData()` and give example on how to complement existing extractions

***
# Before we get started

We first load the birdscanR package.
```{r}
library(birdscanR)
```

Then we set our inputs

Input working folder :
```{r}
# Set main output directory
# =============================================================================
mainOutputDir = file.path(".", "results")
```

Open connection to SQL database : 
```{r}
# Set server and database settings
# =============================================================================
dbServer = "MACHINE\\SERVERNAME" # Set the name of your SQL server
dbName = "db_Name" # Set the name of your database
dbDriverChar = "SQL Server" # Set either "SQL Server" or "PostgreSQL"

# Open the connection with the database
# ===========================================================================
dsn = paste0(
  "driver=", dbDriverChar, ";server=", dbServer,
  ";database=", dbName,
  ";uid=", rstudioapi::askForPassword("Database user"),
  ";pwd=", rstudioapi::askForPassword("Database password")
)
dbConnection = RODBC::odbcDriverConnect(dsn)
```

`ExtractDbData()` will query the SQL database using a sequence of functions (starting with 'get'), calculating day and night information, and compiling these data into a list (`dbData`).

If for some reason, some data were not extracted, or the extraction last too long, it is possible to extract single database tables using the specific SQL-wrapper function.

List of SQL-wrapper functions:
```{r get-functions, echo=TRUE, results='asis'}
get_functions <- grep("^get", getNamespaceExports("birdscanR"), value = TRUE)
# cat("### BirdscanR functions warpping SQL query\n")
cat(paste0("- `", get_functions, "()`"), sep = "\n")
```

Below some example of their utilities outside the main function `extractDbData()`.

***
# Retrieve echoes

Use `getCollectionTable()` to get all echoes registered during 'timeInterval'. Set 'timeInterval = NULL' to extract all registered echoes.
```{r}
# load collection table
# =============================================================================
message("Extracting collection table from DB...")
timeInterval = NULL # e.g. c(as.Date("2024-03-01"),as.Date("2024-03-15"))
echoData = getCollectionTable(dbConnection, dbDriverChar, timeInterval)
```

***
# Retrieve additional echo features

Additional features are stored in an other table. Set a list of features, extract from the database, and merge with the echoes data set.


List of features to extract (link to full list **ToDo : description fo the features now in compileData, but document somewhere as stand-alone yamal ?**
) :
```{r}
# Example: Get wing beat frequency and credibility :
listOfRfFeaturesToExtract = c(167, 168)
```

Extract the features from the database for specific echoes (see `echoData$row`)
```{r}
# Get the requested rf features for the echoes
# =============================================================================
message("Extracting rffeatures table from DB...")
echoRfFeatureMap = getEchoFeatures(dbConnection, dbDriverChar,
  listOfRfFeaturesToExtract = listOfRfFeaturesToExtract,
  echoIDRange = c(
    min(echoData$row),
    max(echoData$row)
  )
)
```

Add the newly extracted features to the existing echoData :
```{r}
echoData %>% left_join(echoData, echoRfFeatureMap, by = join_by(row == echo))
```


***
# Extract information on study site

Key parameters are start and end dates, site location, and for (older) campaign time zone (time shift from UTC).
```{r}
# load site from local MS-SQL DB
# =============================================================================
message("Extracting site table from DB...")
siteTable = getSiteTable(dbConnection, dbDriverChar)
```

