---
title: "create and read VPTS"
author: Birgen Haest, Baptiste Schmid
output: rmarkdown::html_vignette
description: > 
  Test functionality and interoperability for createVPTS
vignette: >
  %\VignetteIndexEntry{Introduction to birdscanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
``` {r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```
***
# Summary

This vignette provides an example of the basic workflow for creatVPTS using:

1.    Create a VPTS using the BirdScan MR1 exemple data

2.    Read the VPTS with getRad(), and visualise with bioRad::

***
# Before we get started

We first install the develop branch of birdscanR from GitHub, and load the package.
```{r, eval=FALSE}
# Install the develop branch of birdscanR from GitHub
# =============================================================================
pak::pak("BirdScanCommunity/birdscanR@develop")
```

```{r}
library(birdscanR)
library(getRad)
# library(bioRad)
```

Read Test BirdScanMR1 data

The package ships with a sample data extract from a Birdscan MR1 radar
(Sempach, Switzerland), already in the format returned by `extractDbData()`.
We load it directly from the package's `extdata` folder.
```{r}
dbData = readRDS(system.file("extdata",
  "CH_Sempach_2024_SEP24_25_DataExtract.rds",
  package = "birdscanR"
))
```

***
# Step 1: Create VPTS from the BirdScan MR1 example dataset

VPTS CSV files, in line with the [ALOFT data standard](https://aloftdata.eu/vpts-csv/),
can be generated directly from the echo data with `createVPTS()`. Note that
this function only works on Birdscan MR1 database versions >= 1.7.0.4, as the
variable `feature37.speed` is required for the density calculation.

```{r}
dbName = "CH_Sempach_2024_SEP24_25"
mainOutputDir = file.path(".", "results")
targetTimeZone = "Etc/GMT0"
timeRangeData = c("2024-09-24 00:00", "2024-09-25 23:59")

# Set manual blind times to NULL (no manual blind times)
# =============================================================================
cManualBlindTimes = NULL

# Create vpts files
# =============================================================================
vptsDir = createVPTS(
  dbName = dbName,
  outputDir = mainOutputDir,
  echoes = dbData$echoData,
  altitudeRange = c(25, 1025),
  altitudeBinSize = 50,
  timeRange = timeRangeData,
  timeBinDuration_sec = 1800,
  timeZone = targetTimeZone,
  protocolData = dbData$protocolData,
  visibilityData = dbData$visibilityData,
  siteData = dbData$siteData,
  sunriseSunset = dbData$sunriseSunset,
  manualBlindTimes = cManualBlindTimes,
  saveBlindTimes = FALSE,
  blindTimesOutputDir = mainOutputDir,
  blindTimeAsMtrZero = NULL,
  propObsTimeCutoff = 0.2
)
```


***
# Step 2: Read the VPTS back in with getRad

`createVPTS()` writes one daily file named `[radarID]_vpts_[yyyyMMdd].csv` into
the `vpts` subdirectory of `outputDir` (i.e. `vptsDir`). The
[getRad](https://github.com/aloftdata/getRad) package can read such local VPTS
CSV files with `get_vpts()` by setting the `path` argument and matching the
local file naming pattern via the `getRad.vpts_local_path_format_aloft`
option.

```{r}
# Tell getRad how the local vpts files are named, so it can find them
# =============================================================================
options(getRad.vpts_local_path_format_aloft = "{radar}_vpts_{year}{month}{day}.csv")

# Read the vpts files created above back in
# =============================================================================
radarID = dbData$siteData$radarID
vpts = getRad::get_vpts(
  radar = radarID,
  datetime = as.Date(c("2024-09-24", "2024-09-25")),
  path = vptsDir
)
vpts
```

```{r}
plot(bioRad::regularize_vpts(vpts))
plot(vpts)
```


